Prepping the Data 
========================================================

## Possible Move Spaces  

For now, we want to get an overall idea of the size of the blocks that each individual can move to. 

```{r}
# How many type 1 seeds do we have? 

how_many_possible<-function( B.new, N, d, field, NumVars, type1seeds, Legal_Index, blocks){
  
  # Checking for possible block moves
  
  SchoolLevel = B.new[,field]
  #SchoolLevel = as.numeric(as.factor(SchoolLevel))
  
  # This part is fixed and fed in 
  check.gap.1 = category.block( c(1,1,1,1,1,1), NumVars, d)
  check.gap.2 = category.block( c(1,1,1,1,1,2), NumVars, d)
  check.gap   = check.gap.2 - check.gap.1
  school.levels = d[field]
  add.vec   = cumsum( rep(check.gap,school.levels-1))
  
  possible.out = rep(0,N)
  
  a = 1 
  
  for( i in 1:N ){
    
    i = i + N 
    
    print(i)
    
    if( a %in% type1seeds){
      possible.out[a] = 0 
    } else{ 
      # which index are we in? 
      x = blocks[i]
      # which level of school? 
      s = which(levels(SchoolLevel)==SchoolLevel[i])
      #SchoolLevel[i] 
    
      up.move   = school.levels - s
      down.move = s -1 
      
      up.check  = x + add.vec[1:up.move]
      if(down.move > 0){
        down.check  = x - add.vec[1:down.move]
      }else{
        down.check = NULL 
      }
      
      # Now, we check to see which of these are legal indices
      
      possible.moves = c(up.check, down.check)
      possible.moves = intersect(possible.moves,Legal_Index)
      possible.moves = length(possible.moves)
      
      possible.out[a] = possible.moves 
    } 
    
    a = a + 1 
    
  }
  
  return(possible.out)
  
  
}

N = nrow(Data)

possoutN = how_many_possible(B,N,d,field=6,NumVars=6,type1Seeds,Legal_Index,blocks)
```

Basically, the above function looks at all the possible school values that we COULD propose for a given record, and examines which of these would be accepted as legal moves. 

If the space of possible moves is small, then we might consider some sort of alternative. I want to try a few ways of doing this and see what happens. 

Note that the above code is structured to only check for moves for File 2 records. We make the assumption in this simulation that File 1 is correct. This is logical, because really we are more concerned about the matches than the correction.  

```{r}
# Which records have a possible block to move to, other than their own? 
nonzero = possoutN[which(possoutN>0)]
# On average, how many possible moves do we have for records that have at least one?
summary( nonzero )
#How many records can move? 
length(nonzero)
#Take a look at the number of moves in a table
table(nonzero)
# Plot
hist(nonzero)
```

The real issue that we see here is the number of possible moves. Some of these records can move to over 70 different blocks. That's a huge number to try and explore. The probabilities at that point also start to get very tiny, so sampling the probability using the DP is going to be potentially problematic. 

We should also consider the records which have no options. 
```{r}

zero.only = which(possoutN==0)
non.only  = which(possoutN>0)
# Some of these are seeds. 
zero.only = setdiff(zero.only,type1Seeds)
length(zero.only)

# Check: These should be the same  
N - n.type1  
sum(length(zero.only)+length(non.only))

which(blocks1[zero.only] != blocks2[zero.only])
# In this case, it means we are off by a field other than school. 

# These records are already in the correct block, and they can't move. Type 2 seeds :) 

Seeds[zero.only] = 2

Data = WholeThing

save.image("AllDataSpace.RData")

```

## Introduce Errors

We need to introduce errors in Ethnicity. 

```{r}
set.seed(9292015)

# How many type 1 seeds?
 n.type1/N
# Roughly 80%

# So, we can only have up to 20% error. Let's try 15%. 

## %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% ##
##     Introduce The Errors           ## 
## %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% ##   

set.seed(104957389)
Data = WholeThing
Data[,"ethnic"]= factor(Data[,"ethnic"])
Data[,"ethnic2011"]= factor(Data[,"ethnic2011"])

Error.Options =  setdiff( 1:N, c(type1Seeds,Race_Nmatch))
error.perc = 15
in.error       = sample( Error.Options, N*error.perc/100, replace = F)
rm(Error.Options)

Data <- within(Data, ethnic<-relevel(ethnic, ref = 6))
Data <- within(Data, ethnic2011<-relevel(ethnic2011, ref = 6))

error.fields = "ethnic"
error.file = 2 
field.options = levels(Data[,error.fields])
fields.probs = table(Data[,"ethnic"])/N

for( i in in.error){
  
  if( error.file == 1 ){
    current.field   = File1[ i ,error.fields]
    field.choices   = setdiff( field.options,current.field )
    File1[i,"prog"] = sample(field.choices,1)
  } else{ 
    current.field   = Data[ i ,error.fields]
    field.choices   = setdiff( field.options,current.field )
    which.field = which(field.options==current.field)
    probs.here = fields.probs[-which.field]
    probs.here = fields.probs[-which.field]/sum(probs.here)
    Data[i,error.fields]=sample(field.choices,1,prob=fields.probs[-which.field])
  }
    
}

sum(Data[,"ethnic2011"]!=Data[,"ethnic"])/N

table(Data[-type1Seeds,"ethnic"])
table(Data[-type1Seeds,"ethnic2011"])
```

## Removing the Superfluous 

At this point, we have things that we do not actually need in this data space. We can remove those. They are stored in "DataSpace.RData" if we need them later. 

```{r}

rm(B12.holder, WholeThing, AllDays11, AllDays12, BDNmatch, BDay2011, BDay2012, BDmatch, Day_Diff, DisagreePool, G.K, Imp.Holder1, Imp.Holder2, Month_Diff, NonEmpty, NumVars, PI, Race_Nmatch, Race_match, SchOnly, SchoolLevel, School_Nmatch, School_match, SeedPool,SeedsOnly, Sex_Nmatch, Singletons, ToRemove, ToRemoveErrors, Year11, Year12, Year_Diff, block.size, check.gap, check.gap.1, check.gap.2, down.move, field, i, inError, k, s, sch11, sch12, school.levels, up.move, x, zero.only, School_Nmatch, School_match, Sex_NMatch, Sex_match, possout, possible.out,small11,small12,smallEither,smallLEA,non.only,nonzero)

save.image("NewErrorData_Unformatted.RData")

```
